National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2024-03-06 analysis/paper/paper.Rmd
Writing in progress.

NOTE: find working version of intro and discussion [here:]
(<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>).

```{=html}
<!--
Motivating example - forecasting impacts of disturbance - WBI and RoF example. In order to reduce uncertainty, need to integrate local information on current state of the population into forecasts.

Lots of variability among caribou ranges in monitoring data: recency, duration, types of data. E.g. 27+ consecutive years of both telemetry-survival and calf/cow-ratios in most Alberta ranges, vs. 1-4 years of data in most Ontario ranges (and that's old) *note we may want to be careful about inter-province comparisons for political reasons, but the variation is super relevant when to-date we've had the same 'rules' apply across the board...

In Missisa (and other places, esp Ontario) available local data are old and limited. Not clear how indicative older data is of state of population. Need new data to resolve questions, but what type, how much is enough. Can't monitor in great detail everywhere all the time. More generally, need for tools to help assess the value of additional monitoring and the best way to invest additional monitoring resources, given the info that is already available. 
-->
```
# Methods

@@

Demographic data for caribou populations, including abundance, vital rates and population growth, are often obtained through the tracking of caribou cows marked with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020]. Although genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021], there is still heavy reliance on collaring. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021]. Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou). Male caribou are not frequently collared.

Collared females are also used in most estimates of recruitment. Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022]. Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below). However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012]. We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details). Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g. @decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. In all scenarios, the projection period begins in 2024. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

@@

We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring varies among scenarios from 1 to 24 years. In all scenarios, monitoring ends in 2023. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range of values to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

TO DO: run low initial population size scenario, put in appendix of one by one sensitivity checks? 
-->
```
## Simulation of local population dynamics and monitoring

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics which is described in the following paragraphs. We refer to this as the "national demographic model" because it is informed by national demographic-disturbance relationships. We use simulated trajectories to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model. For reference, all variables and parameters are defined in supplemental Table A1.

@@

### Simulation of observed survival of collared animals

@@

### Simulation of observed recruitment from composition surveys

@@

The composition survey bias correction term $c$ depends on the known apparent
number of adult females per collared animal $w$, the unknown ratio of young
bulls to adult females $q$, the unknown probability of misidentifying young
bulls as adult females and vice versa $u$, and the unknown probability of
missing a calf $z$ (Supplement B). Prior uncertainty about the value of the bias
correction term $c$ is approximated with a Log-normal distribution
(Supplement B). Given the apparent number of adult females per collared
animal $w$ in each monitoring scenario (Table
\@ref(tab:monitoringScenarios)), we calculate the mean and standard
deviation of 10,000 samples of $\log{c}$ assuming the ratio of young
bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
misidentification probability $u$ varies uniformly between 0 and 0.2,
and the probability of missing a calf $z$ varies uniformly between 0 and
0.2. To maintain consistency with Johnson
et al. [-@johnson_science_2020] we also provide an option to only adjust
for sex ratio and misidentification biases $X_t=cR_t/2$.

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the survival R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

@@

As in the simulations, we assume a simple two-class demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 2000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is whether a
population is declining or not [@dunham_strategic_2023]. If uncertainty about population status
is low, additional investment in reducing uncertainty about the
population growth rate may not be necessary or warranted. Thus, our outcome of interest is how often the Bayesian posterior
mean correctly predicts the true population status (increasing or
decreasing). To calculate this outcome for each disturbance and
monitoring scenario (Fig \@ref(fig:figDisturbance)) we fit the
Bayesian model to simulated observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)). The probability of correct status
assessment is the percentage of these 500 examples in
which the Bayesian posterior mean population growth rate and the true
population growth rate are both greater than 0.99, or both less than or equal to 0.99.

@@

To make this work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios. Where appropropriate, we borrowed and built on code from an app developed by Eacker et al. [-@eacker_web-based_2019]. Our app allows users to provide existing local data or simulate data based on national
models as shown above, define an anthropogenic disturbance scenario and
make projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. The app allows advanced users to adjust many model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g. @dyson_existing_2022; @stewart_climate-informed_2023]. The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when the projection period is longer. Differences are also larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6, and collars are renewed each year. To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Probabilities are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)). When disturbance is very high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases. In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does alter qualitative conclusions (supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

See <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.

@@

Alternatives to aerial composition surveys for estimating recruitment to adult age classes include fecal pellet analysis [@mcfarlane_genetic_2018] and calf collaring [@ellington_divergent_2020]. Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary. We hope that including a bias term in our simple demographic models will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.

@@

Other caribou Bayesian population models have complexities that aren’t included in our simple model, including population size information from aerial or fecal dna surveys, and explicit representation of additional classes [@moeller_integrating_2021; @mcnay_demographic_2022]. At present, our model is intended as a proof-of-concept that can be customized and extended as necessary to meet the needs of particular monitoring programs. In many contexts, population size and fecal dna survey information is not available, and a framework that includes these sources of information will be unnecessarily complicated. Rather than further complicate a proof-of-concept model applied to simulated data, we believe the most effective way forward will be to collaborate with interested decision makers to extend and customize the proof-of-concept model to meet their real local needs, recognizing that these real needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021]. That said, it would also be interesting to compare models with different numbers of age classes to clarify the benefits and drawbacks of that additional complexity.

@@

Other important assumptions of these simple aspatial demographic models are that observed animals are a random and representative sample of a fully mixed population in which survival and recruitment rates apply equally to all animals. However, available genetic evidence suggests high variation in fitness among individuals [@mcfarlane_genetic_2018], and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022]. Risks and benefits for caribou vary spatially [@decesare_linking_2014;@avgar_spaceuse_2015] (more?) and observed variation in habitat preferences and space use among individuals [@avgar_spaceuse_2015] (other habitat selection models with individual variation?) also suggest imperfect mixing and spatial variation in survival and recruitment within populations.

Although there is evidence of spatial structure and high variation in demographic rates among individuals within populations, estimation of spatial variation in demographic rates requires data that is not widely available. That said, DeCesare [-@decesare_linking_2014] successfully estimated spatial variation in adult female survival using 13 years of data from 468 collared animals. Analysis of in metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] can also provide insight into spatial variation in demographic rates. We look forward to a future in which caribou demographic projections are better informed by robust analyses of spatial variation in demographic rates. In the meantime, high variability in demographic rates among individuals [@mcfarlane_genetic_2018] could be incorporated into an aspatial Bayesian population model [@armstrong_incorporating_2021].

In both our Bayesian and simulation models stochastic variation in demographic rates increase when population sizes are very small. In the simulation model, expected recruitment rate is highest at intermediate population sizes, and declines when populations are small or near carrying capacity. Our Bayesian population model does not include this density dependence, and it is not necessary for accurately the viability of our simulated populations (Figure 4). That said, real small populations may face additional challenges such as inbreeding depression and difficulty finding mates [@mills_conservation_2012] that are not included in our models, so the real probability of extinction of very small populations is likely to be higher than projected by these models.

>>>>

NOTE: please provide comments and suggestions for the introduction here:
<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>.

Across 58 study areas for threatened boreal caribou (_Rangifer tarandus caribou_) in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020]. The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011]. Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
Boreal caribou are widely distributed across the Canadian boreal region [@eccc_scientific_2011], inhabiting some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016]. Monitoring rare and widely distributed species such as boreal caribou is expensive, and resources are limited.  Analysis of the potential implications of monitoring outcomes for decisions can clarify that additional monitoring is not always necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012]. A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. For populations of caribou and other species at risk, deciding whether or not management actions will cause declines is important.  Ecological simulation models can be used to evaluate the ability of alternative monitoring options to reduce the risk of errors in species status classifications and population viability projections [@dunham_strategic_2023].

Efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, ignoring an important source of information about outcomes for boreal caribou across their range that can help reduce uncertainty in local demographic projections and status assessments, and limiting their ability to project outcomes on changing landscapes. Existing models have also ignored uncertainty associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012]. 
Here, we present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data and accounts for composition survey errors, building on the work of Eacker et al. (2019). Following the framework proposed by Dunham et al. (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. We also offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data.

# Methods

@@

Caribou demographic data are often obtained by tracking cows with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020]. Although genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021], there is still heavy reliance on collaring. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).

To estimate recruitment, collared females are often located and surveyed by aircraft to estimate the ratio of calves to cows in the groups of animals that include collared animals [@nbckc_nbckc_2022]. Groups typically contain uncollared animals so sample size is typically larger for recruitment than for survival estimates (see multiplier parameter $w$, cowMult below). However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012]. We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details). Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change. In all scenarios, the projection period begins in 2024.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

@@

We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring varies among scenarios from 1 to 24 years. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

TO DO: run low initial population size scenario, put in appendix of one by one sensitivity checks? 
-->
```
## Simulation of population dynamics and monitoring

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics. These example population trajectories are informed by prior information about the distribution of possible parameter values and outcomes from the national analysis of demographic-disturbance relationships [@johnson_science_2020] so we call it the "national" model. Our method for selecting parameter values for each example trajectory yields a continuous set of alternative models (hypotheses) about the state of a particular caribou population, wherein there are more examples of parameter values that prior analysis suggests are more likely; these alternative scenarios informed by prior beliefs are required for analysis of the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the simulated trajectories to confirm that our Bayesian model successfully integrates local data with priors from the national model. For reference, all variables and parameters are defined in supplemental Table A1.

@@

### Simulation of observed survival and recruitment

@@

@@

The composition survey bias correction term $c$ depends on the known apparent
number of adult females per collared animal $w$, the unknown ratio of young
bulls to adult females $q$, the unknown probability of misidentifying young
bulls as adult females and vice versa $u$, and the unknown probability of
missing a calf $z$ (Supplement B). Prior uncertainty about the value of the bias
correction term $c$ is approximated with a Log-normal distribution
(Supplement B). Given the apparent number of adult females per collared
animal $w$ in each monitoring scenario (Table
\@ref(tab:monitoringScenarios)), we calculate the mean and standard
deviation of 10,000 samples of $\log{ c}$ assuming the ratio of young
bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
misidentification probability $u$ varies uniformly between 0 and 0.2,
and the probability of missing a calf $z$ varies uniformly between 0 and
0.2. To maintain consistency with Johnson
et al. [-@johnson_science_2020] we also provide an option to only adjust
for sex ratio and misidentification biases $X_t=cR_t/2$.

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the \ital{survival} R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

@@

As in the simulations, we assume a simple two-class demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 5000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->
We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the \ital{R2jags} R package [@su_r2jags_2021] to fit the Bayesian model. We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2. We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the \ital{mcmcplots} R package [@curtis_mcmcplots_2018].

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the probability of correctly predicting the true population status (increasing or
decreasing) [@dunham_strategic_2023]. To calculate this outcome for each disturbance and
monitoring scenario (Fig \@ref(fig:figDisturbance)) we fit the
Bayesian model to simulated observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), and calculated the percentage of these 500 examples in
which the Bayesian posterior mean population growth rate and the true
population growth rate are both greater than 0.99, or both less than or equal to 0.99.

@@

Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of \ital{tidyverse} packages [@wickham_welcome_2019]. To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al. [-@eacker_web-based_2019] where appropriate. Our app allows users to provide existing local data, or to make projections informed by simulated data as we have done in this analysis (similar to Figure
\@ref(fig:examplePredictions)). The app allows advanced users to adjust many model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, \ital{caribouMetrics}, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g. @dyson_existing_2022; @stewart_climate-informed_2023]. The \ital{caribouMetrics} package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

@@

```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations. In these examples the composition survey cow multiplier $w$ is 6, and collars are renewed each year.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
```

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases. In these examples the composition cow multiplier $w$ is 6; varying $w$ does alter qualitative conclusions (supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

NOTE: Please provide comments and suggestions for the discussion here: <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.

@@

Alternatives to aerial composition surveys for estimating recruitment include fecal pellet analysis [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022] and calf collaring [@ellington_divergent_2020]. Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary. We hope that including a bias term in our simple demographic models will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.

@@

Other caribou Bayesian population models have complexities not included in our simple model, including population size information from aerial or fecal dna surveys, and additional age classes [@moeller_integrating_2021; @mcnay_demographic_2022]. At present, our model is intended as a proof-of-concept that can be customized and extended as necessary to meet the needs of particular monitoring programs. In many contexts, population size and fecal dna survey information is not available, and a framework that includes these sources of information will be unnecessarily complicated. Rather than further complicate a proof-of-concept model applied to simulated data, we believe the most effective way forward will be to collaborate with interested decision makers to extend and customize the proof-of-concept model to meet their real local needs, recognizing that these real needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021]. That said, it would be helpful to clarify the benefits and drawbacks of adding additional age classes.

@@

Other important assumptions of these simple aspatial demographic models are that observed animals are a random and representative sample of a fully mixed population in which survival and recruitment rates apply equally to all animals. However, available genetic evidence suggests high variation in fitness among individuals [@mcfarlane_genetic_2018], and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022]. Risks and benefits for caribou vary spatially [@decesare_linking_2014;@avgar_spaceuse_2015] and observed variation in habitat preferences and space use among individuals [@avgar_spaceuse_2015] (other habitat selection models with individual variation?) also suggest imperfect mixing and spatial variation in survival and recruitment within populations.

Although there is evidence of spatial structure and high variation in demographic rates among individuals within populations, estimation of spatial variation in demographic rates requires data that is not widely available. That said, DeCesare [-@decesare_linking_2014] successfully estimated spatial variation in adult female survival using 13 years of data from 468 collared animals. Metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] can also provide insight into spatial variation in demographic rates. We look forward to a future in which caribou demographic projections are better informed by robust analyses of spatial variation. In the meantime, high variability in demographic rates among individuals [@mcfarlane_genetic_2018] could be incorporated into an aspatial Bayesian population model [@armstrong_incorporating_2021].

In both our Bayesian and simulation models stochastic variation in demographic rates increase when population sizes are very small. Real small populations may face additional challenges such as inbreeding depression and difficulty finding mates [@mills_conservation_2012] that are not included in our models, so the real probability of extinction of very small populations is likely to be higher than projected by these models.
